---
title: "Final Project: Assessing COVID-19 Impacts on Voter Registration in 2020"
author: "Matthew O'Stricker"
embed-resources: true
format: html
editor: visual
echo: false
---

```{r}
#| message: false
library(tidyverse)
library(scales)
library(dplyr)
library(lubridate)
library(broom)
library(moderndive)
theme_set(theme_light())
voter_registration <- read_csv("https://raw.githubusercontent.com/MatthewOStricker/soc223/main/new-voter-registrations.csv")
covid_cases <- read_csv("https://raw.githubusercontent.com/MatthewOStricker/soc223/main/covid-cases.csv")
state_cases <- read_csv("https://raw.githubusercontent.com/MatthewOStricker/soc223/main/Weekly_United_States_COVID-19_Cases_and_Deaths_by_State_-_ARCHIVED_20231109.csv")
us_population <- read_csv("https://raw.githubusercontent.com/MatthewOStricker/soc223/main/united-states-population.csv")
florida_ex <- read_csv("https://raw.githubusercontent.com/MatthewOStricker/soc223/main/Type%20of%20Registration.csv")
fl_tot <- read_csv("https://raw.githubusercontent.com/MatthewOStricker/soc223/main/Voter%20Registration%20Applications%20Received%20by%20Source%20and%20County%202020%20total.csv")
fl_COVID <- read_csv("https://raw.githubusercontent.com/MatthewOStricker/soc223/main/Use%20this%20for%20FL")
fl_pip <- read_csv("https://raw.githubusercontent.com/MatthewOStricker/soc223/main/FL-PIP.csv")

colnames(voter_registration) <- c("State", "Year", "Month", "NewRVoters")

month_order <- c("Jan", "Feb", "Mar", "Apr", "May")

voter_registration$Month <- factor(voter_registration$Month, levels = month_order)

florida_ex$Month <- factor(florida_ex$Month, levels = month_order)

fl_pip$Month <- factor(fl_pip$Month, levels = month_order)

voter_registration <- voter_registration |>
  filter(Month == "Jan" | Month == "Feb" | Month == "Mar" | Month == "Apr" ) 

names(state_cases)[names(state_cases) == "state"] <- "State"

  
state_cases <- state_cases |>
  filter(State == "AZ" | State == "CO" | State == "DE" | State == "DC" | State == "CA" | State == "DC"| State == "FL"| State == "GA"| State == "IL"| State == "MD"| State == "NC"| State == "TX"| State == "VA")


state_abb <- data.frame(
  StateName = state.name,
  StateAbb = state.abb,
  stringAsFactors = FALSE
)

voter_registration$State <- state_abb$StateAbb[match(voter_registration$State, state_abb$StateName)]

voter_registration$State <- ifelse(is.na(voter_registration$State), "DC", voter_registration$State)
```

```{r}
state_cases <- state_cases |>
  mutate(date_updated = mdy(date_updated), 
         start_date = mdy(start_date), 
         end_date = mdy(end_date))
```

```{r}
  state_cases$end_date <- as.Date(state_cases$end_date)
  state_cases$Month <- month(ymd(state_cases$end_date), label = TRUE) 
  
state_cases <- state_cases |>
    filter(Month == "Apr" | Month == "Jan" | Month == "Feb" | Month == "Mar")
```

# 1. Introduction

COVID-19 took the world by storm in 2020 and greatly altered so many key aspects of our daily lives. Outside of the many lives that were taken by the disease, the global pandemic created a shift in the very fabric of American democracy due to the implications it had on the U.S. presidential elections. The U.S. presidential election is a historic staple in American democracy that ultimately decides who will become the leader of the "free world" for the next four years. With the U.S. seeing a constant slight rise in recent years, the expectation is that there would be a constant rise in voter registration in the presidential elections.

To better analyze to see if this occurred and what caused the results that will be found, I plan to answer the research question "What change, if any, occurred to the way voters registered in Florida pre-pandemic - Janurary through February - and post-pandemic - March through April - in 2020?" To answer the question, I plan to use the `voter_registration` dataset from fivethirtyeight and the `covid_cases` dataset that has the covid cases recorded in 2020 in the United States from Our World in Data to analyze the voter registration numbers for each state. To measure the impact, I plan to compare the monthly from Jan-April trends for certain states during the 2020 election cycle to see if I can spot any key differences. I plan to track the change in these months by taking a look at the change in in-person and remote voting during the period.

Since COVID started to gain large traction in the United States in the earlier months of the year 2020, I will be focusing my analysis on Jan-May for the year 2020. The impacts of COVID, which was identified as a global pandemic and passed more than 100,000 deaths in the U.S., were apparent and we should be able to assess what impact, if any, it had on the voter registration in the country.\[\^1\]

Please note, per the CDC, where the COVID-19 cases data was collected, "New York State's reported case and death counts do not include New York City's counts as they separately report nationally notifiable conditions to CDC." This explains why there is info for NY and NYC.

Additionally, only 11 states and Washington D.C. will be included in the datasets because those were the only places where fivethirtyeight was able to get data on new registered voters. I could not find any supplemental data for the other states. However, I believe the 12 jurisdictions will be able to accurately answer the aformentioned question.

In order to conduct effective regression and residual analysis, I plan to use the state of Florida, with a focus on its counties, as a key example for that portion of my final project. Since Florida was the state with the most data accessible, it will serve as the example that can serve as the answer to the research question. I specified my research question to focus on Florida for this reason and hope the answer can potentially serve as the foundation for any future research questions and analysis that focuses on COVID-19's impact other states during 2020.

\[\^1\]: <https://www.usatoday.com/in-depth/news/nation/2020/04/21/coronavirus-updates-how-covid-19-unfolded-u-s-timeline/2990956001/>)

Take a look at the main data I will be analyzing about the counties in Florida during the first four months of 2020:

```{r}
head(fl_pip)
```

# 2. Exploratory Data Analysis

```{r}
ggplot(us_population, aes(x = date, y = Population)) +
  geom_line() +
  geom_point() +
  labs(title = "Population Growth in the United States (2016-2020)",
       x = "Year",
       y = "Population") +
  scale_y_continuous(labels = comma)
```

Figure 1 belows shows that the population in the United States has been growing at a steady rate from 2016 to 2020. The constant growth leads me to hypothesize that there will be a growth in new voter registration numbers due to larger population in the United States. Any dramatic shifts in registration numbers will be alarming and a cause for investigation.

States/Districts Included in my Datasets:

1.  Arizona (AZ)
2.  California (CA)
3.  Colorado (CO)
4.  Washington D.C. (DC)
5.  Delaware (DE)
6.  Florida (FL)
7.  Georgia (GA)
8.  Illinois (IL)
9.  Maryland (MD)
10. North Carolina (NC)
11. Texas (TX)
12. Virginia (VA)

```{r}
  summary_cases <- state_cases |>
    group_by(State, Month) |>
    summarise(
      Mean_cases = mean(new_cases),
      median_cases = median(new_cases),
      SD_cases = sd(new_cases),
    ) |>
    filter(Month <= "Apr") |>
    ungroup()
  
  
  print(summary_cases)
  
```

Figure 1. Population Growth in the United States from 2016 to 2020

After finding the population growth, I wanted to do a dive on the year 2020. First, I graphed Figure 2, which shows the total COVID-19 cases by month for the states included in the datasets, to gauge the amount of cases and their growth trends. It seems the total amount of cases was steadily growing and peaked during the month of March. Since I am able to see there was a large amount of cases during the month, I will want to look at the voter registration during the months to see if I can spot anything of significance. I hypothesize there will be a decline in voter registrations from the month of March to April in 2020.

```{r}

ggplot(state_cases, aes(x = Month, y = tot_cases)) +
  geom_bar(stat = "identity", fill = "red") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  labs(title = "Total COVID-19 Cases by Month in 2020",
       x = "Month",
       y = "Mean COVID Cases") 
```

Figure 2. Total COVID-19 Cases by Month in 2020

```{r}
ggplot(voter_registration, aes(x = Month, y = NewRVoters, fill = as.factor(Year)))+
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  facet_wrap(~Year) +
  labs(title = "Comparing the New Registered Voters in 2016 to 2020 from Jan-May",
       fill = "Year",
       y = "New Registered Voters")
```

Figure 3: Comparing the New Registered Voters from 2016 to 2020 from Jan-May

As we can see from Figure 3, there was quite a shift that occurred. In 2016, there was an expected trend of voter registration steadily increasing for the first three months before experiencing a marginal decline in April. However, in 2020, there was an extremely high peak in February that was followed by a decent drop in March and a massive decline in April. Figure 3 highlights the ever-apparent result: some factor contributed to the drastic decrease in voter registrations from March to April in 2020 that prevented the chart from following the similar trend from 2016. I hypothesize that COVID-19 was a key contributor to the drastic decline in voter registration from March to April in 2020. Now, begins the quest of running analysis to identity any potential correlation between Covid-19 cases and voter registration.

```{r}
state_cases_reduced <- state_cases |>
  group_by(State, Month) |>
  summarise(TotalNewCases = sum(new_cases))
```

```{r}
florida_ex <- florida_ex |>
  pivot_longer(cols = c(InPerson, Remote), names_to = "RegistrationMethod", values_to = "Value")
```

```{r}
ggplot(florida_ex, aes(x = Month, y = Value, color = RegistrationMethod, group = RegistrationMethod)) +
  geom_line(linewidth = 1.5) +
  labs(title = "Comparison of Methods of Voter Registration", y = "Number of Registered Voters", x = "Month") +
  scale_color_manual(name = "Registration Method",
                     values = c("InPerson" = "blue", "Remote" = "red"))
```

Figure 4: Comparison of methods of Voter Registration

One would assume, barring outside circumstances like COVID-19, the number of in-person voters would increase over time at the onset of the year. Figure 4, which answers the key question of how voters register, found a similar area of drastic decline following February through April. For this graph, the sharp decrease in registration was restricted to in-person voters and not remote voters. So, this figure explains that something occurred after February that resulted in a decline in the amount of voters who registered in person. Remote registrants experienced a marginal decline, which means that these voters were not extremely impacted by the outside factor as compared to the in-person voters. Knowing that COVID-19 cases were on the rise at the same time, a thread of correlation between COVID-19 and voter registrants have begun to take shape.

# 3. Multiple linear regression

## 3.1 Methods 

## 3.2 Model Results

## 3.3 Interpreting the regression table

## 3.4 Inference for multiple regression

## 3.5 Residual Analysis

# 4. Discussion

## 4.1 Conclusions

## 4.2 Limitations

## 4.3 Further questions

# 5. Citations and References

```{r}
  fit1 <- lm(pip ~ postcovid * MedianIncome, data = fl_pip)

  get_regression_table(fit1)
```

```{r}
  fit2 <- lm(pip ~ postcovid + MedianIncome, data = fl_pip)

  get_regression_table(fit1)
```

```{r}
ggplot(fl_pip, aes(x = MedianIncome, y = pip, color = factor(postcovid))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Median Income", y = "PIP", color = "Post-COVID", title = "PIP vs. Median Income of Florida Country Registered Voters", subtitle = "Data from Counties in 2020 Pre- and Post-COVID Outbreak")  
```

b2(MedianIncome)

y = b0 + b1(postcovid) + b2(medinc) + b3(postcovid) (MedianInome)

```{r}
pred_data <- augment(fit1)

glimpse(pred_data)

```

```{r}
ggplot(pred_data, aes(x = .resid)) +  
  geom_histogram(binwidth = 7, fill = "blue", color = "white") + 
  labs(title = "Residual Histogram", x = "Residuals") +
  theme_minimal()
```

```{r}
ggplot(pred_data, aes(x = MedianIncome, y = .resid)) +  
  geom_jitter() + 
  geom_smooth(method = lm, se = FALSE, col = "blue") +
  labs(title = "Residuals vs. Median Income", y = "Residuals") +
  theme_minimal()
```

```{r}
ggplot(pred_data, aes(x = MedianIncome, y = .resid)) +  
  geom_boxplot() + 
  geom_smooth(method = lm, se = FALSE, col = "blue") +
  labs(title = "Residuals from Registered Voter Analysis", y = "Residuals") +
  facet_wrap(~postcovid) +
  theme_minimal()
```
